The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-12 manuscript/manuscript.Rnw
discuss RLN

The human gene \textit{CGA} codes for the alpha subunit of multiple
hormones including pituitary hormones (luteinizing hormone,
follicle-stimulating hormone, thyroid-stimulating hormone), as well as
numerous chorionic gonadotropins with variable beta subunits
\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of
the chorionic gonadotropins is to prevent the regression of the corpus
luteum \cite{Hanson.ea1971:EffectsofHCGand}. Among the six placentas
we analyzed here, \textit{CGA} is only expressed in the placentas of
\human and \bonobo (\cref{fig:hormone_genes,fig:cga_evolution_tree}),
which corresponds with the emergence of the chorionic gonadotropins in
anthropoid primates
\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly,
\textit{CGA} is expressed at much higher levels in both species than
the beta subunits with which its product dimerizes (\cref{fig:hormone_genes,fig:cga_evolution_tree,fig:cgb_evolution_tree}).
\textit{CGA} is one of several genes noted by
\citet{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone
adaptive evolution of the \human{}-\textit{P. troglodytes} stem lineage.
including increasing maternal blood flow, widening the pubic bone and
relaxing uterine musculature in preparation for labor. In both dogs
and elephants, there is high expression of relaxin in the term
placenta, with little evidence in other species examined here (Table
8). Relaxin is known to be produced in the placenta of the pregnant
dog and is used in the diagnosis of canine pregnancy
\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof};
however, the source of relaxin produced by pregnant African elephants
remains unclear \cite{Meyer.ea2004:Speciesandfetalgender}. The high
rise in relaxin in elephants has been suggested to be a major signal
for the initiation of labor
\cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive} Here we
show that the placenta of elephants show high expression of relaxin
and is likely the predominant source of relaxin at term \DLA{(Table 8)}.

Two genes, NPPB and NPPC, code for peptide hormones that are important
in the inducement of vasodilation, and may be important in increasing
maternal blood flow to the fetus (Cameron et al., 1996). There is some
indication that these proteins are present in the decidua of mice
(Cameron et al., 1996), but we find very little expression in the
fetal portion of the mouse placenta (Table 8). Here we find high
expression of at least one of these genes in dog, cow, and elephant.
The lack of expression in the hemochorial placentas of
Euarchontoglires may mean that high expression of these genes are
essential in increasing maternal blood flow in epitheliochorial and
endotheliochorial placentas

>>>>

The human gene \textit{CGA} codes for the alpha subunit of multiple
hormones including pituitary hormones (luteinizing hormone,
follicle-stimulating hormone, thyroid-stimulating hormone), as well as
numerous chorionic gonadotropins with variable beta subunits
\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of
the chorionic gonadotropins is to prevent the regression of the corpus
luteum \cite{Hanson.ea1971:EffectsofHCGand}. Among the six placentas
we analyzed here, \textit{CGA} is only expressed in the placentas of
\human and \bonobo (\cref{fig:hormone_genes,fig:cga_evolution_tree}),
which corresponds with the emergence of the chorionic gonadotropins in
anthropoid primates
\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly,
\textit{CGA} is expressed at much higher levels in both species than
the beta subunits with which its product dimerizes (\cref{fig:hormone_genes,fig:cga_evolution_tree,fig:cgb_evolution_tree}).
\textit{CGA} is one of several genes noted by
\citet{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone
adaptive evolution of the \human{}-\textit{P. troglodytes} stem lineage.

\subsubsection{Relaxin}

Relaxin (\textit{RLN}) is a hormone that has multiple effects during
pregnancy including increasing maternal blood flow, widening the pubic
bone and relaxing uterine musculature in preparation for labor. In
\dog, \elephant, \mouse, \carmeli, and \horse there is high expression
of relaxin in the term placenta, with little evidence in other species
examined here (\cref{fig:rln2_evolution_tree}). \textit{RLN} is known
to be produced in the placenta of the pregnant \dog and is used in the
diagnosis of canine pregnancy
\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof},
the luteal ovary and placenta of \pig
\cite{Hisaw.ea1948:Relaxininovaryof}, and the placenta of \horse
\cite{Klonisch.ea1997:Placentallocalizationofrelaxin}. However, the
source of relaxin produced by pregnant \elephant remains unclear
\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in relaxin in
\elephant has been suggested to be a major signal for the initiation
of labor \cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive}.
Here we show that the placenta of elephants show high expression of
relaxin and is likely a major source of \textit{RLN} at term. In
contrast to the results of
\cite{Evans.ea1993:mouserelaxingenenucleotide}, we also see expression
of \textit{RLN} in the placenta of \mouse and \carmeli.

in the inducement of vasodilation, and may be important in increasing
maternal blood flow to the fetus (Cameron et al., 1996). There is some
indication that these proteins are present in the decidua of mice
(Cameron et al., 1996), but we find very little expression in the
fetal portion of the mouse placenta (Table 8). Here we find high
expression of at least one of these genes in dog, cow, and elephant.
The lack of expression in the hemochorial placentas of
Euarchontoglires may mean that high expression of these genes are
essential in increasing maternal blood flow in epitheliochorial and
endotheliochorial placentas
